Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-18 treestats_paper.tex
Updated intro and discussion perf aspects.

As a genetic mutation is passed down across generations,
it distinguishes those genomes that have inherited it from those that have not,
providing a glimpse of the genealogical tree relating the genomes to each other at that site.
Statistical summaries of genetic variation therefore also describe the underlying genealogies.
We exploit this correspondence here to define a general framework that
efficiently compute single-site population genetic statistics
using the succinct tree sequence encoding of genealogies and genome sequence.
The general approach accumulates ``sample weights'' within each marginal tree,
which are then combined using a ``summary function'';
different statistics result from different choices of weight and function.
Results can be reported in two dual ways:
by \emph{site}, which corresponds to statistics calculated from genome
sequence; and by \emph{branch}, which gives the expected value of the site statistic
under the infinite-sites model.
We use the framework to implement many currently-defined statistics of genome
sequence (making the statistics' relationship to the underlying genealogical trees
concrete and explicit) and suggest several new statistics.
We evaluate computational performance using simulated data, and
show that calculating statistics from tree sequences using this general
framework is several orders of magnitude more efficient than
optimized matrix-based methods in terms of both run time and memory requirements.
We also explore how well the duality between site and branch statistics holds
in practice on trees inferred from the 1000 Genomes Project dataset,
and discuss ways in which deviations may encode interesting biological
signals.

@@

It was once a major undertaking to collect data sufficient to estimate a single summary
of the many genetic relationships between the individuals of a sample,
e.g., to estimate levels of sequence diversity from a few dozen alleles at an allozyme locus
\citep{Kreitman1983-xr}.
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes \citep[e.g,][]{booker2018understanding,haenel2018metaanalysis,stankowski2019widespread},
or between individuals rather than populations \citep[e.g.,][]{browning2010highresolution,speed2014relatedness,aguillon2017deconstructing}.
Computation is beginning to be a major problem:
genetic data on the scale of the UK Biobank~\citep{bycroft2018genome},
for instance, might have 500K individuals genotyped at 10M loci,
resulting in a genotype matrix of $5 \times 10^{12}$ entries.
Computing a single summary from this matrix is still relatively quick,
once loaded into memory,
but for many applications we are now interested in computing a large number of statistics,
or iteratively refining a single computation.
Furthermore, even larger datasets are being collected,
and millions of whole genomes will soon be available. In this case the
genotype matrix simply cannot fit into memory, and complex distributed computations
are required.

Fortunately, the genotype matrix is not the only way to encode genetic data.
The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity) is a data
structure that efficiently encodes the genealogical trees describing how a set of
individuals are related to each other at each point along their genome.
It was introduced in the context of coalescent
simulation~\citep{kelleher2016efficient}, and lead to scalability increases of
several orders of magnitude over existing methods. The methods were subsequently extended
and refined for forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}, with similarly large efficiency gains.
Recent work has shown that tree sequence algorithms can
also be used to massively increase the scalability of methods for inferring
genome-wide genealogies, and making it possible to infer trees for millions of
samples~\citep{kelleher2019inferring}.
The key to the remarkable efficiency of tree sequence
algorithms is the way that shared structure in adjacent trees along the genome is encoded.
As one looks across the genome, genealogies change at recombination breakpoints in ancestors.
However, nearby trees tend to share a lot of common structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
This redundancy can be exploited to store genome sequence and the
associated genealogical relationships very compactly, while also
enabling efficient calculation of several population genetics
statistics~\citep{kelleher2016efficient}.
This paper generalizes the approach to a much broader class of statistics,
supporting arbitrary tree topologies and patterns of allelic variation.
% Want to emphasise generality here: people might assume we're only dealing
% with the biallelic case.

@@

% overview: general definition that is mostly functions of jAFS but includes others
In this paper we have described a general framework for summarizing genetic variation
and the underlying genealogies
that encompasses many standard population genetic statistics.
Many of these statistics are functions of the joint allele frequency spectrum,
but the framework is more general,
and can be used, for example, to quantify associations between genotypes and phenotypes.
This generality greatly reduces the software development effort in
implementing statistics efficiently,
and it also allows users to easily explore new classes of statistics.
The range of statistics available to describe variation in a single exchangeable sample
(i.e., a single population) are well-understood,
but there are much larger and less well-explored classes of statistics
describing genetic variation between many populations
or across geographical space.
The statistics defined here are all additive along the genome:
if we have computed a statistic in equal-size windows along a chromosome,
then the value of the statistic for the entire chromosome is equal to
the average of the values in those windows.
Some commonly-used statistics (e.g., $F_{ST}$ or Tajima's $D$)
are not additive, but are ratios of additive statistics, so can be easily computed in this way.
% We have also defined site statistics to be additive across \emph{alleles},
% to allow the definitions to apply to sites with more than two alleles.
% This decision has some nice properties, but is not the only reasonable choice.
% Multilocus statistics that measure linkage disequilibrium or haplotype lengths
% do not fall in this category.
Extensions to statistics involving the pairwise joint distribution of genotypes across
sites~\citep{hudson2001twolocus} such as linkage disequilibrium are planned for future work.
Haplotype-based statistics may require different classes of algorithms.

@@

% speed/efficiency: big data sets; large simulations
The most obvious application of these methods on current practice
is to improve efficiency of existing pipelines,
as tree sequences allow storage and processing of large genomic data sets
with orders of magnitude less space and time than most other methods.
All statistics described here (and more) are implemented
in the rigorously tested \tskit{} library,
which provides a suite tools for efficiently working with tree sequences in Python and C.
The largest efficiency gains are found for very large sample sizes
only likely to be available in real data sets for humans in the near future,
but the ability to analyze millions of whole genomes
is useful for understanding evolutionary simulations of any species
\citep[see, e.g.,][]{martin2017human,browning2018one,galloway2019stickleback}.
% TODO rephrase this: large scale simulations are useful for lots of things.
% No need for the slightly negative note about large datasets only being
% relevant for humans, I think.

% TODO: too negative.
% We hope a more broadly interesting and useful aspect of this paper
% is a more in-depth look at the correspondence between statistics calculated with genome sequence
% and properties of the underlying genealogies, i.e.,
% the duality between Site and Branch statistics under the infinite-sites model.
The correspondence between genome sequence and properties of the
underlying genealogies we have used here is well-known. The general framework
that we have defined, however, codifies this relationship in a
mathematically elegant and computationally efficient way, and may lead to
new perspectives on well-studied problems.
One way to use the duality between Site and Branch statistics is to answer:
what aspect of tree shape is a particular population genetics statistic summarizing?
This can help when interpreting results, especially if reality may not fit an idealized model
of separate populations.
However, methods to infer tree sequences from real data
make it now possible to work in the other direction:
instead of calculating (Site) statistics from the genotype matrix,
we can calculate precisely analogous (Branch) statistics from the trees themselves,
thus hopefully bypassing the extra layer of noise induced by mutation.
How well this works will depend on the ability of inference methods
to estimate the true tree sequence.
This might plausibly produce less noisy estimates because tree sequence inference
should use the signal from nearby patterns of variation
to interpret relationships at a given site,
thus transforming the simple binary split induced by a SNP
into a much richer source of information.
Furthermore,
if tree sequence inference can be made insensitive to local variation in mutation rate,
calculation of branch statistics would provide a summary method
that is not affected by this potentially important confounding factor.
Similarly, if tree sequences inferred from
genotype array data~\citep{kelleher2019inferring} are unbiased,
then Branch statistics could provide a way
to estimate genome-wide quantities without ascertainment bias.
This procedure would be similar to imputation, and would likely face similar challenges.

>>>>

As a genetic mutation is passed down across generations,
it distinguishes those genomes that have inherited it from those that have not,
providing a glimpse of the genealogical tree relating the genomes to each other at that site.
Statistical summaries of genetic variation therefore also describe the underlying genealogies.
We use this correspondence to define a general framework that
efficiently computes single-site population genetic statistics
using the succinct tree sequence encoding of genealogies and genome sequence.
The general approach accumulates ``sample weights'' within each marginal tree,
which are then combined using a ``summary function'';
different statistics result from different choices of weight and function.
Results can be reported in two dual ways:
by \emph{site}, which corresponds to statistics calculated from genome
sequence; and by \emph{branch}, which gives the expected value of the site statistic
under the infinite-sites model.
We use the framework to implement many currently-defined statistics of genome
sequence (making the statistics' relationship to the underlying genealogical trees
concrete and explicit) and suggest several new statistics.
We evaluate computational performance using simulated data, and
show that calculating statistics from tree sequences using this general
framework is several orders of magnitude more efficient than
optimized matrix-based methods in terms of both run time and memory requirements.
We also explore how well the duality between site and branch statistics holds
in practice on trees inferred from the 1000 Genomes Project dataset,
and discuss ways in which deviations may encode interesting biological
signals.

@@

It was once a major undertaking to collect data sufficient to estimate a single summary
of the genetic relationships between the individuals of a
sample~\citep[e.g.][]{Kreitman1983-xr}.
% e.g., to estimate levels of sequence diversity from a few dozen alleles at an allozyme locus
% \citep{Kreitman1983-xr}.
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes, both between
individuals~\citep[e.g.,][]{browning2010highresolution,speed2014relatedness,aguillon2017deconstructing},
and within
populations~\citep[e.g,][]{booker2018understanding,haenel2018metaanalysis,stankowski2019widespread}.
Computation is beginning to be a major problem:
projects on the scale of the UK Biobank~\citep{bycroft2018genome}
and gnomAD~\citep{karczewski2019variation} hold genetic data for
hundreds of thousands of samples at tens to hundreds of millions of variant sites.
Such large genotype matrices are extremely unwieldy and difficult to process.
Tools such as \texttt{hail} (\url{https://hail.is}) allow computations to be sharded
across many machines in parallel, thereby calculating statistics far more
quickly than is possible using single-computer methods such as
\texttt{plink}~\citep{purcell2007plink}. Computer time must still be
paid for, however, and running calculations concurrently on thousands of cores quickly
becomes expensive. Larger datasets still, consisting of millions of whole
genomes, are currently being collected, and it is clear that processing
these genotype matrices will be very costly.

The genotype matrix, fortunately, is not the only way to encode genetic data.
The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity) is a data
structure that efficiently encodes the genealogical trees describing how a set of
individuals are related to each other at each point along their genome.
It was introduced in the context of coalescent
simulation~\citep{kelleher2016efficient}, and lead to scalability increases of
several orders of magnitude over existing methods. The methods were subsequently extended
and refined for forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}, with similarly large efficiency gains.
Recent work has shown that tree sequence algorithms can
also be used to massively increase the scalability of methods for inferring
genome-wide genealogies, and making it possible to infer trees for millions of
samples~\citep{kelleher2019inferring}.
The key to the remarkable efficiency of tree sequence
algorithms is the way that shared structure in adjacent trees along the genome is encoded.
As one looks across the genome, genealogies change at recombination breakpoints in ancestors.
However, nearby trees tend to share a lot of common structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
This redundancy can be exploited to store genome sequence and the
associated genealogical relationships very compactly, while also
enabling efficient calculation of several population genetics
statistics~\citep{kelleher2016efficient}.
This paper generalizes the approach to a much broader class of statistics,
supporting arbitrary tree topologies and patterns of allelic variation.
% Want to emphasise generality here: people might assume we're only dealing
% with the biallelic case.

@@

% overview: general definition that is mostly functions of jAFS but includes others
In this paper we have described a general framework for summarizing genetic variation
and the underlying genealogies
that encompasses many standard population genetic statistics.
Many of these statistics are functions of the joint allele frequency spectrum,
but the framework is more general and can be used,
for example, to quantify associations between genotypes and phenotypes.
This generality greatly reduces the software development effort in
implementing statistics efficiently,
and it also allows users to easily explore new classes of statistics.
The range of statistics available to describe variation in a single exchangeable sample
(i.e., a single population) are well-understood,
but there are much larger and less well-explored classes of statistics
describing genetic variation between many populations
or across geographical space.
The statistics defined here are all additive along the genome:
if we have computed a statistic in equal-size windows along a chromosome,
then the value of the statistic for the entire chromosome is equal to
the average of the values in those windows.
Some commonly-used statistics (e.g., $F_{ST}$ or Tajima's $D$)
are not additive, but are ratios of additive statistics, so can be easily computed in this way.
% We have also defined site statistics to be additive across \emph{alleles},
% to allow the definitions to apply to sites with more than two alleles.
% This decision has some nice properties, but is not the only reasonable choice.
% Multilocus statistics that measure linkage disequilibrium or haplotype lengths
% do not fall in this category.
Extensions to statistics involving the pairwise joint distribution of genotypes across
sites~\citep{hudson2001twolocus} such as linkage disequilibrium are planned for future work.
Haplotype-based statistics may require different classes of algorithms.

@@

% speed/efficiency: big data sets; large simulations
The most obvious application of these methods on current practice
is to improve efficiency of existing pipelines,
as tree sequences allow storage and processing of large genomic data sets
with orders of magnitude less space and time than standard matrix-based methods.
All statistics described here (and more) are implemented
in the rigorously tested \tskit{} library,
which provides a suite tools for efficiently working with tree sequences in Python and C.
Large-scale simulations are useful in many contexts
\citep[e.g.,][]{martin2017human,browning2018one,galloway2019stickleback}
and the ability to quickly compute a wide range of statistics from
these (previously prohibitively large) datasets opens up new possibilities.
At smaller scales, the statistics are still highly efficient and avoiding
these cost of exporting data to a genotype matrix will in practice greatly
speed up applications such as deep learning \jk{cite} and ABC~\jk{cite}
that compute many statistics from simulated data. The ability to
efficiently compute statistics from real data is also welcome, of course.

The correspondence between genome sequence and properties of the
underlying genealogies we have used here is well-known. The general framework
that we have defined, however, codifies this relationship in a
mathematically elegant and computationally efficient way, and may lead to
new perspectives on well-studied problems.
One way to use the duality between Site and Branch statistics is to answer:
what aspect of tree shape is a particular population genetics statistic summarizing?
This can help when interpreting results, especially if reality may not fit an idealized model
of separate populations.
However, methods to infer tree sequences from real data
make it now possible to work in the other direction:
instead of calculating (Site) statistics from the genotype matrix,
we can calculate precisely analogous (Branch) statistics from the trees themselves,
thus hopefully bypassing the extra layer of noise induced by mutation.
How well this works will depend on the ability of inference methods
to estimate the true tree sequence.
This might plausibly produce less noisy estimates because tree sequence inference
should use the signal from nearby patterns of variation
to interpret relationships at a given site,
thus transforming the simple binary split induced by a SNP
into a much richer source of information.
Furthermore,
if tree sequence inference can be made insensitive to local variation in mutation rate,
calculation of branch statistics would provide a summary method
that is not affected by this potentially important confounding factor.
Similarly, if tree sequences inferred from
genotype array data~\citep{kelleher2019inferring} are unbiased,
then Branch statistics could provide a way
to estimate genome-wide quantities without ascertainment bias.
This procedure would be similar to imputation, and would likely face similar challenges.
